Discrete Quantum Theories and Computing

### 2018-03-01 dissertation.tex
dissertation.lyx "2.1.1 The single qubit problem" Updating...

A convenience for probability calculations and a necessity for computing
relative state properties is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $\rme^{\rmi\phi}$, that is, 
\begin{equation}
\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\phi}\alpha_{0},\rme^{\rmi\phi}\alpha_{1}\right)\,.\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration \cite{Hatcher2001,MosseriDandoloff2001,Bengtsson2007},
which we can write down as follows: let 
\begin{align}
\alpha_{0} & =x_{0}+\rmi y_{0}\,, & \alpha_{1} & =x_{1}+\rmi y_{1}\,.\label{A4.eq}
\end{align}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

We can reduce 3 real degrees of freedom in \Eq{A5.eq} to 2 real
degrees of freedom by effectively removing $\rme^{\rmi\phi}$ (``fibering
out by the circle $\Sphere{1}$''). The standard form of this maps
(``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\,.\nonumber 
\end{eqnarray}
By denoting the the three-dimensional vector~$\left(X,Y,Z\right)$
as $\Vec{X}$, \Eq{A5.eq} implies these transformed coordinates
obeying 
\begin{equation}
\|\Vec{X}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. In Fig.~\ref{hopfFibers.fig},
we illustrate schematically the family of circles \textit{each one
of which is collapsed to a point\/} $(\theta,\phi)$ on the surface
$X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

%% (We try to list the sources of all computed figures)%% Sources for hopfFibers.fig:  dqc-geom-2012/00mathFigs/Hopf-Fiber.nb
\begin{figure}[htb]
\centerline{ \includegraphics[width=0.48\columnwidth]{hopfS2}\hspace{0.03in}
\includegraphics[width=0.48\columnwidth]{hopfFibers}} \centerline{\hspace{1in}
(a) \hfill{}(b) \hspace{1in}} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration as a family
of circles (the paths of $\rme^{\rmi\phi}$), each corresponding to
a single point on the sphere in (a). Points in (a) are color coded
corresponding to circles in (b), e.g., one pole contains the red elliptical
circle that would become an infinite-radius circle in a slightly different
projection, and the opposite pole corresponds to the large perfectly
round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{eqnarray}
\hspace*{-0.5cm}\left(x_{0},y_{0},x_{1},y_{1}\right) & = & \left(\cos\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\sin\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\ \right.\nonumber \\
 &  & \left.\cos\frac{\theta-\phi}{2}\sin\frac{\psi}{2},\ \sin\frac{\theta-\phi}{2}\sin\frac{\psi}{2}\right),\label{A8a.eq}
\end{eqnarray}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\pi$, with $0\leq\frac{\theta+\phi}{2}<2\pi$ and $0\leq\frac{\theta-\phi}{2}<2\pi$,
we see that 
\begin{eqnarray}
\left(X,Y,Z\right)=\left(\cos\phi\sin\psi,\,\sin\phi\sin\psi,\,\cos\psi\right)\ .\label{A8b.eq}
\end{eqnarray}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to 
\begin{equation}
\ket{\psi_{1}}=\rme^{\rmi\phi}\cos\frac{\psi}{2}\ket{0}+\sin\frac{\psi}{2}\ket{1}\ ,\label{A9.eq}
\end{equation}
and an irreducible state can be represented as a point on a sphere,
as shown in Figure \ref{blochSphere.fig}(a).

>>>>

A convenience for computing probability and relative state properties
is the normalization condition 
\begin{equation}
\|\psi_{1}\|^{2}=|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=\alpha_{0}^{*}\alpha_{0}+\alpha_{1}^{*}\alpha_{1}=1\ ,\label{A2.eq}
\end{equation}
which identifies $\alpha_{0}$ and $\alpha_{1}$ as (complex) probability
amplitudes and implies the conservation of probability in the closed
world spanned by $\{\ket{0},\ket{1}\}$. Note that we distinguish
for future use the \textit{norm\/} $\|\cdot\|$ of a vector from
the \textit{modulus\/} $|\cdot|$ of a complex number. Continuing,
we see that if we want only the irreducible state descriptions, we
must supplement the process of computing \Eq{A2.eq} by finding a
way to remove the distinction between states that differ only by an
overall phase transformation $\rme^{\rmi\phi}$, that is, 
\begin{equation}
\left(\alpha_{0},\alpha_{1}\right)\sim\left(\rme^{\rmi\phi}\alpha_{0},\rme^{\rmi\phi}\alpha_{1}\right)\,.\label{A3.eq}
\end{equation}
This can be accomplished by the Hopf fibration \cite{Hatcher2001,MosseriDandoloff2001,Bengtsson2007},
which we can write down as follows: let 
\begin{align}
\alpha_{0} & =x_{0}+\rmi y_{0}\,, & \alpha_{1} & =x_{1}+\rmi y_{1}\,.\label{A4.eq}
\end{align}
Then \Eq{A2.eq} becomes the condition that the four real variables
describing a qubit denote a point on the three-sphere $\Sphere{3}$
(a 3-manifold) embedded in $\R^{4}$: 
\begin{equation}
x_{0}^{\ 2}+y_{0}^{\ 2}+x_{1}^{\ 2}+y_{1}^{\ 2}=1\ .\label{A5.eq}
\end{equation}

We can reduce 3 real degrees of freedom in \Eq{A5.eq} to 2 real
degrees of freedom by effectively removing $\rme^{\rmi\phi}$ (``fibering
out by the circle $\Sphere{1}$''). The standard form of this maps
(``the Hopf fibration'') is 
\begin{eqnarray}
X & = & 2\,\mathrm{Re}\ \alpha_{0}\alpha_{1}^{*}=2x_{0}x_{1}+2y_{0}y_{1}\,,\nonumber \\
Y & = & 2\,\mathrm{Im}\ \alpha_{0}\alpha_{1}^{*}=2x_{1}y_{0}-2x_{0}y_{1}\,,\\
Z & = & |\alpha_{0}|^{2}-|\alpha_{1}|^{2}=x_{0}^{\ 2}+y_{0}^{\ 2}-x_{1}^{\ 2}-y_{1}^{\ 2}\,.\nonumber 
\end{eqnarray}
By denoting the the three-dimensional vector~$\left(X,Y,Z\right)$
as $\Vec{X}$, \Eq{A5.eq} implies these transformed coordinates
obeying 
\begin{equation}
\|\Vec{X}\|^{2}=X^{2}+Y^{2}+Z^{2}=\left(|\alpha_{0}|^{2}+|\alpha_{1}|^{2}\right)^{2}=1\label{A7.eq}
\end{equation}
and therefore have only two remaining degrees of freedom describing
all possible distinct one-qubit quantum states. 
\begin{figure}[tbh]
\centerline{ \includegraphics[width=0.48\columnwidth]{hopfS2}\hspace{0.03in}
\includegraphics[width=0.48\columnwidth]{hopfFibers}} \centerline{\hspace{1in}
(a) \hfill{}(b) \hspace{1in}} \caption{(a) The sphere represented by \Eq{A7.eq}, which is the irreducible
space of one-qubit states, along with a representative set of points
on the sphere. (b) Representation of the Hopf fibration as a family
of circles (the paths of $\rme^{\rmi\phi}$), each corresponding to
a single point on the sphere in (a). Points in (a) are color coded
corresponding to circles in (b), e.g., one pole contains the red elliptical
circle that would become an infinite-radius circle in a slightly different
projection, and the opposite pole corresponds to the large perfectly
round red circle at the equator.}
\label{hopfFibers.fig} 
\end{figure}
In Fig.~\ref{hopfFibers.fig}, we illustrate schematically the family
of circles \textit{each one of which is collapsed to a point\/} $(\theta,\phi)$
on the surface $X^{2}+Y^{2}+Z^{2}=1$ by the Hopf map.

The resulting manifold is the two-sphere $\Sphere{2}$ (a 2-manifold)
embedded in $\R^{3}$. If we choose one of many possible coordinate
systems describing $\Sphere{3}$ via \Eq{A5.eq} such as 
\begin{equation}
\left(x_{0},y_{0},x_{1},y_{1}\right)=\left(\cos\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\,\sin\frac{\theta+\phi}{2}\cos\frac{\psi}{2},\ \cos\frac{\theta-\phi}{2}\sin\frac{\psi}{2},\ \sin\frac{\theta-\phi}{2}\sin\frac{\psi}{2}\right)\,,\label{A8a.eq}
\end{equation}
where %% $0\leq \theta \leq 2 \pi$,  $-\pi \leq \phi \leq  \pi$, 
$0\leq\psi\leq\pi$, with $0\leq\frac{\theta+\phi}{2}<2\pi$ and $0\leq\frac{\theta-\phi}{2}<2\pi$,
we see that 
\begin{equation}
\left(X,Y,Z\right)=\left(\cos\phi\sin\psi,\,\sin\phi\sin\psi,\,\cos\psi\right)\,.\label{A8b.eq}
\end{equation}
Thus the one-qubit state is independent of $\theta$, and we can choose
$\theta=\phi$ without loss of generality, reducing the form of the
unique one-qubit states to 
\begin{equation}
\ket{\psi_{1}}=\rme^{\rmi\phi}\cos\frac{\psi}{2}\ket{0}+\sin\frac{\psi}{2}\ket{1}\,,\label{A9.eq}
\end{equation}
and an irreducible state can be represented as a point on a sphere,
as shown in Figure \ref{blochSphere.fig}(a).
